---
title: "DIABLO WHONDRS"
author: "Flores, Javier E & Degnan, David J"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(mixOmics) 
library(tidyverse)
library(data.table)
library(kneedle) # devtools::install_github("etam4260/kneedle")
library(tidymodels)
library(fastshap)
library(shapviz)
library(tidymodels)

seed = 2825
set.seed(seed)
```

## Format Multiomics Data

DIABLO assumes samples are rows and columns are the features, and requires a list
of data.frames. 

```{r}
transpose_df <- function(x) {
  x %>%
    pivot_longer(2:ncol(.)) %>% 
    rename(Sample = name) %>% 
    pivot_wider(names_from = Feature, id_cols = Sample) %>%
    dplyr::select(-Sample) %>%
    as.matrix()
}

# Load datasets 
metab = fread("../../Dataset/Scaled/Metabolomics.txt")
lip_pos = fread("../../Dataset/Scaled/Lipidomics_Positive.txt")
lip_neg = fread("../../Dataset/Scaled/Lipidomics_Negative.txt")

# Make list of input matrices
omicsX = list(
  "metabolomics" = transpose_df(metab), 
  "lipidomics positive" = transpose_df(lip_pos),
  "lipidomics negative" = transpose_df(lip_neg)
)

# Make response data.frame
responseY = data.frame(
  group = factor(colnames(metab)[-1] %>% substr(1,1), levels = c("N", "S"))
)
```

## Fine Tune DIABLO Model

Due to DIABLO's design, it is difficult to use the exact same CV partitions as we
used for the other models. 

*Create Design Matrix*

```{r}
# Fit PLS between all pairs of data in omicsX. 
pls_res <- list()
for(i in 1:(length(omicsX)-1)){
  for(j in (i+1):length(omicsX)){
    chari <- as.character(i)
    charj <- as.character(j)
    pls_res[[chari]][[charj]] <- spls(X = omicsX[[i]], 
                                      Y = omicsX[[j]], 
                                      ncomp = 2,
                                      # Note that in fitting spls, one must specify the 
                                      # number of features with non-zero loadings there should be in each of the
                                      # final components. Choice of this number is fairly arbitrary, given that
                                      # the algo works to find linear combinations that maximize the covariance 
                                      # between the indicated datasets. Therefore, the features with non-zero loadings
                                      # will represent those that are *most* correlated between the two datasets. 
                                      keepX = rep(34, 2), # Cannot be any larger than the smallest ome - so 34
                                      keepY = rep(34, 2)) # Cannot be any larger than the smallest ome - so 34
    rm(chari, charj)
  }
}

dat_cors <- expand.grid(1:length(omicsX), 1:length(omicsX)) %>%
  as.data.frame(.) %>%
  dplyr::filter(Var1 != Var2) %>%
  dplyr::filter(Var1 < Var2) %>%
  dplyr::mutate(correlation = NA,
                design_weight = 0)
design <- matrix(0, ncol = length(omicsX), nrow = length(omicsX), # for square matrix filled with 0.1s
                 dimnames = list(names(omicsX), names(omicsX)))
for(i in 1:nrow(dat_cors)){
  chari <- as.character(dat_cors[i,1])
  charj <- as.character(dat_cors[i,2])
  dat_cors$correlation[i] <- cor(pls_res[[chari]][[charj]]$variates$X, 
                                 pls_res[[chari]][[charj]]$variates$Y)[1,1]
  # Here is where threshold is implemented.
  dat_cors$design_weight[i] <- ifelse(dat_cors$correlation[i] > 0.75, 0.1, 0)
  design[dat_cors[i,1], dat_cors[i,2]] <- dat_cors$design_weight[i]
  design[dat_cors[i,2], dat_cors[i,1]] <- dat_cors$design_weight[i]
  rm(chari,charj)
}
# Add col/row corresponding to Y linkage. Values here should always be 1.
design <- rbind(cbind(design, 1), 1)
diag(design) <- 0
rownames(design) = colnames(design) = c(colnames(design)[-(length(omicsX)+1)], "Y")

design
```

*Tune the number of components and features in a component*

```{r}
## Select error type by design
#if(all(diff(table(responseY)) == 0)){
#  
#  # Choice if balanced design
#  error_choice <- "Overall.ER"
#  measure <- "overall" # used in next tuning step for tune.block.splsda
#  
#} else{
#  
#  # Choice if unbalanced design
#  error_choice <- "Overall.BER"
#  measure <- "BER" # used in next tuning step for tune.block.splsda
#}
#
## Let's do 10%, 25%, 50%, and 100%
#test.keepX <- list(metabolomics = round(c(0.1, 0.25, 0.5, 1)*nrow(metab)),
#                   `lipidomics positive` = round(c(0.1, 0.25, 0.5, 1)*nrow(lip_pos)),
#                   `lipidomics negative` = round(c(0.1, 0.25, 0.5, 1)*nrow(lip_neg)))
#
## Maximum number of components to consider - based on the maximum value before error outs
#test.ncomp <- 7
#
## All values are default unless otherwise indicated
#BPPARAM <- BiocParallel::SnowParam(workers = parallel::detectCores()/4)
#system.time({
#  tune.model <- tune.block.splsda(X = omicsX, # non-default
#                                  Y = responseY$group, # non-default  
#                                  ncomp = test.ncomp, # non-default, got error using initial amount
#                                  # "Error in solve.default(Sr): system is computationally singular: reciprocal condition number = 3.22428e-19" 
#                                  # hence the substraction
#                                  test.keepX = test.keepX, # non-default
#                                  design = design, # non-default
#                                  measure = measure, # non-default
#                                  validation = 'Mfold', # non-default
#                                  folds = 3, # non-default
#                                  nrepeat = 5, # non-default
#                                  dist = "mahalanobis.dist", # non-default
#                                  weighted = TRUE,
#                                  progressBar = TRUE,
#                                  tol = 1e-6,
#                                  max.iter = 100, 
#                                  near.zero.var = FALSE,
#                                  #scheme = "horst",
#                                  scale = TRUE, # Scaling already occurs
#                                  #init = "svd",
#                                  light.output = TRUE,
#                                  signif.threshold = 0.01,
#                                  BPPARAM = BPPARAM, # non-default
#                                  seed = 1) # non-default
#})
#BiocParallel::bpstop()
#
## Extract the optimal number of components
#final.ncomp <- tune.model$choice.ncomp$ncomp
#
## Extract the optimal number of features that should have non-zero loadings
## for each dataset. Need to reduce down to vector sets based on final.ncomp
#final.keepX <- lapply(tune.model$choice.keepX, function(x, ncomps){
#  x[1:ncomps]
#}, ncomps = final.ncomp)
```

Apply final tuning parameters and save model. 

```{r}
final.model <- block.splsda(X = omicsX, # non-default
                            Y = responseY$group, # non-default  
                            ncomp = 1,
                            keepX = list(metabolomics = 54, 
                                         `lipidomics positive` = 8,
                                         `lipidomics negative` = 17),
                            design = design,
                            #scheme = "horst", # This is the default value, even if doc doesn't show it
                            scale = TRUE,
                            #init = "svd",
                            tol = 1e-06,
                            max.iter = 100,
                            near.zero.var = FALSE,
                            all.outputs = TRUE,
                            verbose.call = FALSE)
```

Build final model.

```{r}
final_mod_predinfo <- predict(object = final.model, newdata = omicsX)
```

## Build U matrix 

Note that here, there is one u matrix per view. Here, let's take an average.

```{r}
u_mat = do.call(cbind, final.model$variates) %>%
  data.frame() %>%
  mutate(
    group = ifelse(grepl("N", colnames(metab)[-1]), "None", "Surrounding"),
    Factor1 = rowMeans(.),
  ) %>%
  dplyr::select(group, Factor1) 

fwrite(u_mat, "../../Comparison/u_matrix/DIABLO_umat_average.csv")

u_mat
```

```{r}
u_mat %>%
  mutate(Factor2 = NA) %>%
  ggplot(aes(x = Factor1, y = Factor2, color = group)) +
    geom_point(size = 2)
```

## Construct V Matrix

```{r}
# Construct V Martix - Factor, Weight, Feature, View
v_mat <- do.call(rbind, final.model$loadings[1:3]) %>%
  data.frame() %>%
  mutate(
    Factor = "Factor1",
    View = lapply(1:(length(final.model$loadings)-1), function(el) {
      rep(names(final.model$loadings)[el], nrow(final.model$loadings[[el]]))
    }) %>% unlist(),
    Feature = row.names(.)
  ) %>%
  rename(Weight = comp1) %>%
  dplyr::select(Factor, Weight, Feature, View)
row.names(v_mat) <- 1:nrow(v_mat)

# Write out the v matrix
fwrite(v_mat, "../../Comparison/v_matrix/DIABLO_vmat.csv", quote = F, row.names = F)

v_mat
```

## Detect Top Features with Kneedle

```{r}
kneedle_df <- v_mat %>%
  dplyr::select(Weight, View, Feature) %>%
  mutate(`Absolute Weight` = abs(Weight)) %>%
  arrange(-`Absolute Weight`) %>%
  mutate(Rank = 1:nrow(.)) 
knee <- kneedle(kneedle_df$Rank, kneedle_df$`Absolute Weight`)

fwrite(kneedle_df %>% dplyr::select(View, Feature, `Absolute Weight`, Rank), 
       "../../Comparison/kneedle_df/DIABLO_kneedle.csv", quote = F, row.names = F)

# Let's plot
ggplot(kneedle_df, aes(x = Rank, y = `Absolute Weight`, color = View)) +
  geom_point() +
  theme_bw() +
  geom_hline(yintercept = knee[2], linetype = "dashed") +
  ggtitle("DIABLO") +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom")
```

```{r}
kneedle_df %>%
  filter(Rank <= knee[1]) %>%
  dplyr::select(View, Feature, `Absolute Weight`, Rank) %>%
  fwrite("../../Comparison/TopK/DIABLO_TopK.csv", quote = F, row.names = F)
```

```{r}
kneedle_df %>%
  filter(Rank <= knee[1]) %>%
  dplyr::select(View, Feature, `Absolute Weight`, Rank)
```



